An Analysis of Performance Bottlenecks in MRI Pre-Processing

### 2024-02-13 paper/paper.tex
[WIP]

\usepackage{cite}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\usepackage{graphicx}
\graphicspath{{figures}}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup[subfigure]{singlelinecheck=false, skip=0pt}
\usepackage{listings}
\renewcommand{\lstlistingname}{Algorithm}
\usepackage[binary-units=true]{siunitx}
\usepackage{csvsimple}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{import}

@@

% DL
In recent years, Deep Learning (DL) has shown promising results in MRI pre-preprocessing. For example, tools such as SynthMorph~\cite{Hoffmann2022-hu}, SynthSeg~\cite{Billot2023-vp}, SynthSR~\cite{Iglesias2023-bp}, and SynthStrip~\cite{Hoopes2022-ms} were integrated into FreeSurfer~\cite{Fischl2012-cx} with the effort to provide Deep Learning alternative to traditional algorithms. The main benefit of those tools is the faster runtime they provide. Moreover, A study~\cite{Pepe2023-dm} showed that the results generated by SynthMorph are numerical more stable than those from FreeSurfer. However, training a DL model requires large amount of data and specialized hardware, and is computationally expensive. Additionally, the results from DL models can be challenging to interpret, explaining the slow adoption of DL in MRI pre-processing.

@@

% - Pipelines
Neuroscientists developed various toolboxes to tackle the challenging task of pre-processing MRI data. We focus on the commonly accepted fMRIPrep~\cite{Esteban2019-bl} workflow as it a comprehensive pre-processing pipeline for both sMRI and fMRI. fMRIPrep combines several pipelines to produce a complete pre-processing workflow. We also study its sub-workflow separately to provide a finer grain analysis of the performance bottleneck.

@@

% - VTune profiler
To profile the pipelines, we use the Intel VTune profiler~\cite{vtune-profiler}. The VTune profiler is a multi-language profiler with low performance overhead and provides information of functions at runtime level. We profile the application using two threading configuration: single-threaded and 32 threads. We aggregate subjects' results for each pipeline to obtain an average performance profile. We compiled each application with debug information to obtain human readable information on function and module names in VTune.

\section{Materials \& Methods}
\subsection{Pipelines}
In this paper, we study the performance bottleneck of fMRIPrep; a commonly used fMRI and sMRI pre-processing workflow. First, we profile the entire workflow to identify the coarse grain performance bottleneck to obtain a coarse view of the workflow's performance. Following, to characterize the performance at a finer grain, we profile the sub-workflows of fMRIPrep: ANTs brainExtraction, ANTs registrationSyN~\cite{Avants2008-ea}, FreeSurfer recon-all~\cite{Dale1999-wu}, FSL FAST~\cite{Zhang2001-hx}, FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}, and FSL MCFLIRT~\cite{Jenkinson2002-od}. We omitted some sub-workflows of fMRIPrep as they were not compatible with our dataset. Moreover, the statistical analysis of the fMRIPrep workflow are not computationally expensive and are not the focus of this paper.

\subsubsection{ANTS registrationSyN}
% Spatial Normalization
% SyN Registration: https://doi.org/10.1016/j.media.2007.06.004
% TODO Mention the option for single and double precision.

\subsubsection{ANTS brainExtraction}
% ANTS brainExtraction
% 1. Truncate image intensity
% 2. N4 bias correction: https://doi.org/10.1109/TMI.2010.2046908
% 3. SyN Registration: : https://doi.org/10.1016/j.media.2007.06.004
% 4. Warp prior to image using affine
% 5. Atropos: https://pubmed.ncbi.nlm.nih.gov/21373993/
% TODO Mention the option for single and double precision.
% TODO OASIS template citation

\subsubsection{FSL FAST}
% DOI: https://doi.org/10.1109/42.906424
% Segmentation

\subsubsection{FreeSurfer recon-all}
% Surface Reconstruction: https://doi.org/10.1006/nimg.1998.0395
% Refer to the fMRIPrep specific 3 stages workflow.

\subsubsection{FSL FLIRT}
% Registration
% FSL FLIRT: https://doi.org/10.1006/nimg.2002.1132, https://doi.org/10.1016/S1361-8415(01)00036-6, https://doi.org/10.1016/j.neuroimage.2009.06.060

\subsubsection{FSL MCFLIRT}
% Head-Motion Correction
% FSL MCFLIRT: https://doi.org/10.1006/nimg.2002.1132

\subsubsection{fMRIPrep}
\MD{TODO import fMRIPrep citation boilerplate from "./fmriprep-citations"}

@@

The Intel VTune profiler solves these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level. However, common profiling challenges remain. Profilers require source code to be compiled with Debug Symbols to report useful information about the different functions and modules. Moreover, different input data can substantially vary the analysis due to conditional branching in the application and convergence thresholds. We discuss our approach to these challenges in the next section.
\MD{TODO Cite effect from varying input data on numerical stability and reduced precision. Ask yohan for refereces}

First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image. For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}} \lstlistingname~\ref{alg:vtune_example} illustrate the profiling of a pipeline by mounting the VTune binary to the Apptainer image at execution.

\begin{minipage}{\linewidth}
	\lstinputlisting[
		language={sh},
		caption={Profiling an application within a Apptainer container, using VTune profiler.},
		captionpos=b,
		label=alg:vtune_example,
		numbers=left,
		frame=single,
		basicstyle=\footnotesize,
		numbersep=5pt,
		numberstyle=\tiny\color{gray},
		rulecolor=\color{black},
		tabsize=2,
	]{vtune-example.sh}
\end{minipage}

Second, we aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile. The Pareto principle is a power law which states that 80\% of the effects come from 20\% of the causes. We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines, which is consistent with the literature~\cite{Kukunas2015-jd}. Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the application. This heuristic allows us to focus our efforts on the performance critical sections of a pipeline. We report the mean and standard deviation of the execution time for those functions.

@@

\begin{figure*}[h!]
	\noindent
	\begin{minipage}
		{0.5\textwidth}
		\begin{align*}
			C & = CPU\_CLK\_UNHALTED.THREAD                \\
			S & = CYCLE\_ACTIVITY.STALLS\_LDM\_PENDING     \\
			O & = CYCLE\_ACTIVITY.STALLS\_L1D\_PENDING     \\
			T & = CYCLE\_ACTIVITY.STALLS\_L2\_PENDING      \\
			W & = MEM\_L3\_WEIGHT                          \\
			H & = MEM\_LOAD\_UOPS\_RETIRED.LLC\_HIT        \\
			R & = MEM\_LOAD\_UOPS\_MISC\_RETIRED.LLC\_MISS \\
			M & = \frac{W \times R}{H + R}                 \\
			N & = \frac{H}{H + W \times R}                 
		\end{align*}
	\end{minipage}
	\begin{minipage}
		{0.5\textwidth}
		\begin{align}
			\%~Memory~Bound & = \frac{S}{C} \times 100          \\
			\%~L1~Bound     & = \frac{S - O}{C} \times 100      \\
			\%~L2~Bound     & = \frac{O - T}{C} \times 100      \\
			\%~L3~Bound     & = \frac{T \times N}{C} \times 100 \\
			\%~DRAM~Bound   & = \frac{T \times M}{C} \times 100 
		\end{align}
	\end{minipage}
	\caption{Intel performance monitoring events and derived memory metrics}
	\label{fig:memory-metrics}
\end{figure*}

The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load. L1 bound occurs when a demand load stall some CPU cores although the data already reside in L1 cache. Similarly, L2, L3, and DRAM bound occurs when a demand load stall the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively. The store bound occurs when a store operation stall the CPU. Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory boudnd characterization~\cite{Intel2006-lc}. We use the convention from~\cite{Kukunas2015-jd} for the equations.

@@

@@

\section{Results}
In this section we present the aggregated profiling data obtained with VTune. First, we present a performance overview of the pipelines. We then dive into more detailed analysis observed from the hotspots of the pipelines. Unless specified explicitly, we only present the profile data from using 1-thread since the results from 32-threads are similar.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{figures/global-longtail.png}
	\caption{Distribution of the functions' CPU time. The left y-axis shows the average total CPU Time spent in a function, while the right y-axis show the cumulative CPU time percentage. The x-axis percentage of functions ordered by decreasing CPU time. The data includes all functions from all pipelines.}
	\label{fig:long-tail-distribution}
\end{figure}

\subsection{Makespan: A performance overview}
Figure~\ref{fig:makespan-1thread} depicts the total time elapsed from the start to the end of each pipeline; i.e. makspan. There is a significant difference between the makespan of different applications. This is expected as the pipelines perform different tasks. The standard deviation is nearly null for all pipelines, except for FreeSurfer recon-all which as a small variation. Overall, this shows that the input data does not substantially affect the makespan of those pipelines.

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
	\caption{Average makespan of the pipeline while using 1-thread. The error bars show the standard deviation. \TG{Font size in the figure is too small, it should be similar to the caption font size}}
	\label{fig:makespan-1thread}
\end{figure}

@@

\csvnames{csvcol}{1=\pipeline, 2=\mem, 3=\la, 4=\lb, 5=\lc, 6=\dram, 7=\store}
\csvstyle{memory_bound}{
	tabular = |r|c|c|c|c|c|c|,
	table head = \hline Pipeline &  \% Memory Bound &  \% L1 Bound &  \% L2 Bound &  \% L3 Bound & \% DRAM Bound & \% Store Bound\\\hline\hline,
	late after line = \\\hline,
	respect all,
	csvcol
}
\begin{table*}[ht]
	\centering
	\csvreader[memory_bound]{tables/memory_bound.csv}{}
	{\pipeline & \tablenum[round-precision=2]{\mem} & \tablenum[round-precision=2]{\la} & \tablenum[round-precision=2]{\lb} & \tablenum[round-precision=2]{\lc} & \tablenum[round-precision=2]{\dram} & \tablenum[round-precision=2]{\store}}
	\caption{Impact of data load on stalled CPU cycles. The values are the summation of the metric weighted by each function CPU time. This represent the percentage of the total CPU time stalled by each metrics. We show the average value across all subjects execution with one thread.}
	\label{extab:memory-bound}
\end{table*}

\subsection{Main Bottleneck: Linear Interpolation}
Table~\ref{extab:interpolation} shows that interpolation (or convolution) is a critical part for all pipelines, except FreeSurfer recon-all. For those pipelines, interpolation contributes between 32.20\% to 62.70\% of the total CPU time. Overall, the number of functions using interpolation is low with less than 20 in each applications. In other words, few interpolation functions are contribute to a large portion of the CPU time.
% There seems to be few  the stages from recon-all: https://surfer.nmr.mgh.harvard.edu/fswiki/recon-all (see section "Manual-Intervention Workflow Directives": stages 1-31)

@@

\subsection{ANTS: Single vs. Double Precision}
Figure~\ref{fig:makespan-ants} shows the makespan of ANTS brainExtraction and ANTS registrationSyN in double and single precision. For both pipelines, the makespan is significantly higher in single precision. This is unexpected, as the floating-point arithmetic operations are supposed to be faster in single precision~\cite{Wang2018-jv}.

\begin{figure}[h]
	\includegraphics[width=\linewidth]{figures/makespan-ants.png}
	\caption{Comparison of makespan between double (blue) and single (orange) precision for ANTS brainExtraction and ANTS registrationSyN.}
	\label{fig:makespan-ants}
\end{figure}

The number of iteration between both version of ANTS registrationSyN is similar. In fact, for the latest two stages, they were the same across all but one execution. However, The single precision version of ANTS registrationSyN took longer per iteration than the double precision version (Figure~\ref{fig:mean-time-per-iteration-ants}). Therefore, the slowdown is not due to the convergence of the algorithm, but rather the execution time for each iteration.

\begin{figure}
	\includegraphics[width=\linewidth]{figures/ants-registrationSyN-iteration-mean.png}
	\caption{Mean time per iteration for ANTS registrationSyN in double and single precision. Only the SyN Registration stage is shown as the two earlier stage were near zero time. The error bars show the standard deviation.}
	\label{fig:mean-time-per-iteration-ants}
\end{figure}

\subsection{FreeSurfer: Thread-Synchronization}
Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant different in the pipeline profile between single-threaded and multi-threaded execution. The multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP is a major bottleneck for the application, accounting for at least 76\% of the application runtime. FreeSurfer uses the multi-processing API provided by OpenMP. Moreover, the y-axis scale is multi folds larger than for the single-threaded executions. This indicates that the multi-threaded execution is significantly slower than the single-threaded execution.

@@

%%%%%%%%%% # Linear Interpolation
% Theoretical improvements using reduced precision. If time permits, toy experiment with ITK single precision data type at different locations.
% 1. Winograd algorithm for fast convolution.
% - It could be used to speed up the current pipeline. Tradeoff with precision? How does it compare to reduced precision?
% 2. Improvement to interpolation would help the current algorithms, as well as the DL algorithm using CNN.
\subsection{Importance of Linear Interpolation}
\MD{TODO}

%%%%%%%%%% # Single vs. Double Precision
% 1. Not as simple as using different data type. It could affect
% - Convergence
% - Accuracy
% - Runtime (maybe even slower)
% 1.5 Convergence is the sane in our case (same number of iterations).
% 2. Why is it slower in our case?
% -? Possibly less optimized code for single precision.
% -? Possibly memory placement
% -? ...
% 3. 
\subsection{ANTS: A tale of precision}
\MD{TODO + change title}

\subsection{OpenMP bottleneck in FreeSurfer recon-all}
In our experiments, FreeSurfer recon-all showed a significant decrease in parallel efficiency with an increase in the number of threads. While the Amdhals' law plays an impact on the maximum efficiency, the OpenMP thread synchronization played the largest impact in our results. The codebase use 83 \textit{static} scheduling policy out of 92 total. \textit{Static} scheduling assign chunks to threads in round-robin fashion. This could lead to a load imbalance between the threads, and thus a decrease in parallel efficiency. We supposed that \textit{dynamic} scheduling could improve the parallel efficiency, by assigning chunks based on the current load of the threads. Unfortunately, we failed to recompile FreeSurfer by naively changing the \textit{static} OpenMP scheduling to \textit{dynamic}. This remain challenging as the codebase is large and complex, and the change in scheduling policy could lead to unexpected bugs.

Future work should study the different configuration of OpenMP scheduling in FreeSurfer recon-all to improve the parallel efficiency. We believe that significant performance improvements could be achieved by optimizing the thread synchronization. This would lead to faster runtime time and higher CPU usage ration in HPC allocations.

@@

Our analysis omits the space or time of execution for function calls. This information could provide additional insight for optimization. For example, spatial and temporal dimensions were shown to be important in the field of reduced precision. Future work could extent the depth of the analysis by including this information.
\MD{TODO Cite paper on spatial/temporal for RP (ask yohan for references)}

@@

\section{Data Availability}
\label{sec:data-availability}
The entire OpenNeuro ds004513 v1.0.2 dataset is freely available to download at:
\\\href{https://openneuro.org/datasets/ds004513/versions/1.0.2}{https://openneuro.org/datasets/ds004513/versions/1.0.2}.

The container images for our profiling experiments are publicly on Docker Hub:
\begin{itemize}
	\item mathdugre/cmake:debug-info
	\item mathdugre/intel-compilers:debug-info
	\item mathdugre/afni:debug-info
	\item mathdugre/ants:debug-info
	\item mathdugre/fsl:debug-info
	\item mathdugre/freesurfer:debug-info
	\item mathdugre/fmriprep:debug-info
\end{itemize}

Our profiling results from VTune are publicly available at:
\\\href{URL}{\MD{TODO add Zenodo link}}

\section{Code Availability}
\label{sec:code-availability}
The code to compile, profile the pipelines, and generate figures is publicly available at:
\\\href{https://github.com/mathdugre/mri-bottleneck}{https://github.com/mathdugre/mri-bottleneck}
\MD{TODO Update URL to slashbin org, when created.}

\section*{Acknowledgement}
\MD{TODO}
% NSERC PGS-D
% Concordia Bursary (find name)
% Any from others?
% ...

@@

\newpage
\onecolumn
\section*{Supplementary Materials}
\begin{figure*}
	\centering
	\includegraphics{figures/global-zipf_law.png}
	\caption{Distribution of the functions' CPU time compare to a Zipf's law distribution with $\alpha=1.8$. The y-axis shows the average CPU time for a function. The x-axis shows the percentage of functions ordered by decreasing CPU time. The data includes all functions from all pipelines.}
	\label{sup-fig:zips-law}
\end{figure*}

@@

\csvnames{csvcol}{2=\module, 3=\func, 4=\mean, 5=\std}
\csvstyle{hotspot}{
	tabular = |r|l|l|c|,
	table head = \hline & Module & Function & CPU Time (mean$\pm$std)\\\hline\hline,
	late after line = \\\hline,
	respect all,
	csvcol
}
\newcommand{\csvtable}[3]{
	\begin{table}[h]
		\resizebox*{\textwidth}{!}{
			\csvreader[hotspot]{#1}{}
			% \csvreader[hotspot, filter={\value{csvrow}<60}]{#1}{}
			{\thecsvrow & \module & \func & \tablenum[round-precision=2, round-mode=places]{\mean}$\pm$\tablenum[round-precision=2, round-mode=places]{\std}}
		}
		\caption{#2}
		\label{#3}
	\end{table}
}
% CSV tables
% \csvtable{tables/hotspots-1thread-freesurfer-reconall.csv}{FreeSurfer recon-all (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-freesurfer-reconall}
\csvtable{tables/hotspots-32threads-freesurfer-reconall.csv}{FreeSurfer recon-all (32 threads): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-32threads-freesurfer-reconall}

>>>>

\usepackage{cite}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\usepackage{graphicx}
\graphicspath{{figures}}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup[subfigure]{singlelinecheck=false, skip=0pt}
\usepackage{listings}
\renewcommand{\lstlistingname}{Algorithm}
\usepackage[binary-units=true]{siunitx}
\usepackage{csvsimple}
\usepackage{booktabs}
\usepackage{amsmath}

@@

% DL
% SynthMorph, SynthSeg, etc.
% Not broadly adopted yet
\MD{TODO Deep Learning approach}

@@

% - Pipelines
Neuroscientists developed various toolboxes to tackle the challenging task of pre-processing MRI data. We focus on the commonly accepted fMRIPrep~\cite{Esteban2019-og} workflow as it a comprehensive pre-processing pipeline for both sMRI and fMRI. fMRIPrep combines several pipelines to produce a complete pre-processing workflow. We also study the sub-workflows of fMRIPrep: ANTs brainExtraction, ANTs registrationSyN, FreeSurfer recon-all, FSL FAST, FSL FLIRT, and FSL MCFLIRT.
\MD{TODO Cite the sub-workflows}

@@

% - VTune profiler
To profile the pipelines, we use the Intel VTune profiler~\cite{vtune_profiler}. The VTune profiler is a multi-language profiler with low performance overhead and provides information of functions at runtime level. We profile the application using two threading configuration: single-threaded and 32 threads. We aggregate subjects' results for each pipeline to obtain an average performance profile. We compiled each application with debug information to obtain human readable information on function and module names in VTune.

% - Outline of the paper.
\MD{TODO outline?}

\section{Materials \& Methods}
\subsection{Pipelines}
In this paper, we study the performance bottleneck of fMRIPrep~\cite{Esteban2019-og}; a commonly used fMRI and sMRI pre-processing workflow. First, we profile the entire workflow to identify the coarse grain performance bottleneck to obtain a coarse view of the workflow's performance. Following, to characterize the performance at a finer grain, we profile the sub-workflows of fMRIPrep: ANTS brainExtraction, ANTS registrationSyN, FreeSurfer recon-all, FSL FAST, FSL FLIRT, and FSL MCFLIRT. We omitted some sub-workflows of fMRIPrep as they were not compatible with our dataset. Moreover, the statistical analysis of the fMRIPrep workflow are not computationally expensive and are not the focus of this paper.

\subsubsection{ANTS registrationSyN}
% Spatial Normalization
% SyN Registration: https://pubmed.ncbi.nlm.nih.gov/17659998/
% TODO Mention the option for single and double precision.

\subsubsection{ANTS brainExtraction}
% ANTS brainExtraction
% 1. Truncate image intensity
% 2. N4 bias correction: https://ieeexplore.ieee.org/document/5445030
% 3. SyN Registration: : https://pubmed.ncbi.nlm.nih.gov/17659998/
% 4. Warp prior to image using affine
% 5. Atropos: https://pubmed.ncbi.nlm.nih.gov/21373993/
% TODO Mention the option for single and double precision.

\subsubsection{FSL FAST}
% Segmentation

\subsubsection{FreeSurfer recon-all}
% Surface Reconstruction
% Refer to the fMRIPrep specific 3 stages workflow.

\subsubsection{FSL FLIRT}
% Registration

\subsubsection{FSL MCFLIRT}
% Head-Motion Correction
% Refer to FSL FLIRT

@@

The Intel VTune profiler~\cite{vtune_profiler} solves these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level. However, common profiling challenges remain. Profilers require source code to be compiled with Debug Symbols to report useful information about the different functions and modules. Moreover, different input data can substantially vary the analysis due to conditional branching in the application and convergence thresholds. We discuss our approach to these challenges in the next section.
% TODO add citation on effect of different input data on numerical stability and reduced precision.

First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image. For use on HPC systems, we created Singularity images using Docker2Singularity \lstlistingname~\ref{vtune_example} illustrate the profiling of a pipeline by mounting the VTune binary to the Singularity image at execution.
% TODO make sure the algo span one page.
\lstinputlisting[
	language={sh},
	caption={Profiling an application within a Singularity container, using VTune profiler.},
	captionpos=b,
	label=vtune_example,
	numbers=left,
	frame=single,
	basicstyle=\footnotesize,
	numbersep=5pt,
	numberstyle=\tiny\color{gray},
	rulecolor=\color{black},
	tabsize=2,
]{vtune-example.sh}

Second, we aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile. The Pareto principle is a power law which states that 80\% of the effects come from 20\% of the causes. We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines. Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the application. This heuristic allows us to focus our efforts on the performance critical sections of a pipeline. We report the mean and standard deviation of the execution time for those functions.

@@

% cpu_clk_unhalted.thread: [Core cycles when the thread is not in halt state]
% cycle_activity.stalls_ldm_pending: [Execution stalls while memory subsystem has an outstanding load]
% cycle_activity.stalls_l1d_pending: [Execution stalls while L1 cache miss demand load is outstanding]
% cycle_activity.stalls_l2_pending: [Execution stalls while L2 cache miss demand load is outstanding]
% MEM_L3_WEIGHT: Approximately the external memory to L3 cache latency ratio
% MEM_LOAD_UOPS_RETIRED.LLC_HIT: [Retired load uops which data sources were data hits in L3 without snoops required Supports address when precise. Spec update: BDM100 (Precise event)]
% MEM_LOAD_UOPS_MISC_RETIRED.LLC_MISS: [Miss in last-level (L3) cache. Excludes Unknown data-source Supports address when precise. Spec update: BDM100, BDE70 (Precise event)]
The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load. L1 bound occurs when a demand load stall some CPU cores although the data already reside in L1 cache. Similarly, L2, L3, and DRAM bound occurs when a demand load stall the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively. The store bound occurs when a store operation stall the CPU. Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive those metrics.
\MD{TODO cite intel reference manual and "Intro to Profiling" ch. 6. Should I cite here or in the figure caption?}

@@

\begin{figure*}
	\noindent
	\begin{minipage}
		{0.5\textwidth}
		\begin{align*}
			C & = CPU\_CLK\_UNHALTED.THREAD                \\
			S & = CYCLE\_ACTIVITY.STALLS\_LDM\_PENDING     \\
			O & = CYCLE\_ACTIVITY.STALLS\_L1D\_PENDING     \\
			T & = CYCLE\_ACTIVITY.STALLS\_L2\_PENDING      \\
			W & = MEM\_L3\_WEIGHT                          \\
			H & = MEM\_LOAD\_UOPS\_RETIRED.LLC\_HIT        \\
			R & = MEM\_LOAD\_UOPS\_MISC\_RETIRED.LLC\_MISS \\
			M & = \frac{W \times R}{H + R}                 \\
			N & = \frac{H}{H + W \times R}                 
		\end{align*}
	\end{minipage}
	\begin{minipage}
		{0.5\textwidth}
		\begin{align}
			\%~Memory~Bound & = \frac{S}{C} \times 100          \\
			\%~L1~Bound     & = \frac{S - O}{C} \times 100      \\
			\%~L2~Bound     & = \frac{O - T}{C} \times 100      \\
			\%~L3~Bound     & = \frac{T \times N}{C} \times 100 \\
			\%~DRAM~Bound   & = \frac{T \times M}{C} \times 100 
		\end{align}
	\end{minipage}
	\caption{Intel performance monitoring events and derived memory metrics~\MD{TODO cite intel reference manual and "Intro to Profiling" ch. 6}.}
	\label{fig:memory-metrics}
\end{figure*}

@@

\section{Results}
\MD{TODO Find better titles for the subsections.}
In this section we present the aggregated profiling data obtained with VTune. First, we present a performance overview of the pipelines. We then dive into more detailed analysis observed from the hotspots of the pipelines. Unless specified explicitly, we only present the profile data from using 1-thread since the results from 32-threads are similar.

@@

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\linewidth]{figures/global-longtail.png}
	\caption{Distribution of the functions' CPU time. The left y-axis shows the average total CPU Time spent in a function, while the right y-axis show the cumulative CPU time percentage. The x-axis percentage of functions ordered by decreasing CPU time. The data includes all functions from all pipelines.}
	\label{fig:long-tail-distribution}
\end{figure}

\subsection{Makespan: A performance overview}
Figure~\ref{fig:makespan-1thread} depicts the total time elapsed from the start to the end of each pipeline; i.e. makspan. There is a significant difference between the makespan of different applications. This is expected as the pipelines perform different tasks. The standard deviation is nearly null for all pipelines, except for FreeSurfer recon-all which as a small variation. Overall, this shows that the input data does not substantially affect the makespan of those pipelines.
\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
	\caption{Average makespan of the pipeline while using 1-thread. The error bars show the standard deviation. \TG{Font size in the figure is too small, it should be similar to the caption font size}}
	\label{fig:makespan-1thread}
\end{figure}

@@

\csvnames{csvcol}{1=\pipeline, 2=\mem, 3=\la, 4=\lb, 5=\lc, 6=\dram, 7=\store}
\csvstyle{memory_bound}{
	tabular = |r|c|c|c|c|c|c|,
	table head = \hline Pipeline &  \% Memory Bound &  \% L1 Bound &  \% L2 Bound &  \% L3 Bound & \% DRAM Bound & \% Store Bound\\\hline\hline,
	late after line = \\\hline,
	respect all,
	csvcol
}
\begin{table*}[h]
	\centering
	\csvreader[memory_bound]{tables/memory_bound.csv}{}
	{\pipeline & \tablenum[round-precision=2]{\mem} & \tablenum[round-precision=2]{\la} & \tablenum[round-precision=2]{\lb} & \tablenum[round-precision=2]{\lc} & \tablenum[round-precision=2]{\dram} & \tablenum[round-precision=2]{\store}}
	\caption{Impact of data load on stalled CPU cycles. The values are the summation of the metric weighted by each function CPU time. This represent the percentage of the total CPU time stalled by each metrics. We show the average value across all subjects execution with one thread.}
	\label{extab:memory-bound}
\end{table*}

\subsection{Linear Interpolation is the Main Bottleneck}
Table~\ref{extab:interpolation} shows that interpolation (or convolution) is a critical part for all pipelines, except FreeSurfer recon-all. For those pipelines, interpolation contributes between 32.20\% to 62.70\% of the total CPU time. Overall, the number of functions using interpolation is low with less than 20 in each applications. In other words, few interpolation functions are contribute to a large portion of the CPU time.

\MD{TODO Check why FreeSurfer recon-all does not use interpolation. Possibly hidden in other functions or using different algorithm.}

@@

\subsection{Single vs. Double Precision}
Figure~\ref{fig:makespan-ants} shows the makespan of ANTS brainExtraction and ANTS registrationSyN in double and single precision. For both pipelines, the makespan is significantly higher in single precision. This is unexpected, as the floating-point arithmetic operations are supposed to be faster in single precision \MD{TODO Cite quadratic of bit-witdh}.
\begin{figure}[h]
	\includegraphics[width=\linewidth]{figures/makespan-ants.png}
	\caption{Comparison of makespan between double (blue) and single (orange) precision for ANTS brainExtraction and ANTS registrationSyN.}
	\label{fig:makespan-ants}
\end{figure}

\MD{TODO Discuss the potential reasons for the slowdown. The difference in time for iterations in single and double precision.}
The number of iteration between both version of ANTS registrationSyN is similar. In fact, for the latest two stages, they were the same across all but one execution. However, The single precision version of ANTS registrationSyN took longer per iteration than the double precision version (Figure~\ref{fig:mean-time-per-iteration-ants}). Therefore, the slowdown is not due to the convergence of the algorithm, but rather the execution time for each iteration.
\begin{figure}
	\includegraphics[width=\linewidth]{figures/ants-registrationSyN-iteration-mean.png}
	\caption{Mean time per iteration for ANTS registrationSyN in double and single precision. Only the SyN Registration stage is shown as the two earlier stage were near zero time. The error bars show the standard deviation.}
	\label{fig:mean-time-per-iteration-ants}
\end{figure}

\subsection{Thread-Synchronization using OpenMP}
Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant different in the pipeline profile between single-threaded and multi-threaded execution. The multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP is a major bottleneck for the application, accounting for at least 76\% of the application runtime. FreeSurfer uses the multi-processing API provided by OpenMP. Moreover, the y-axis scale is multi folds larger than for the single-threaded executions. This indicates that the multi-threaded execution is significantly slower than the single-threaded execution.
\MD{TODO Mention the OpenMP scheduling used in majority.}

@@

\MD{TODO}
%%%%%%%%%% # Linear Interpolation
% Theoretical improvements using reduced precision. If time permits, toy experiment with ITK single precision data type at different locations.
% 1. Winograd algorithm for fast convolution.
% - It could be used to speed up the current pipeline. Tradeoff with precision? How does it compare to reduced precision?
% 2. Improvement to interpolation would help the current algorithms, as well as the DL algorithm using CNN.

%%%%%%%%%% # Single vs. Double Precision
% 1. Not as simple as using different data type. It could affect
% - Convergence
% - Accuracy
% - Runtime (maybe even slower)
% 1.5 Convergence is the sane in our case (same number of iterations).
% 2. Why is it slower in our case?
% -? Possibly less optimized code for single precision.
% -? Possibly memory placement
% -? ...
% 3.

%%%%%%%%%% OpenOMP bottleneck (FreeSurfer with 32 threads)
% Hypothesis: OpenMP scheduling is not optimal for the pipeline.
% - What are the potential issues with the current scheduling policy?
% Future work: Modify the OpenMP scheduling to improve the parallel efficiency.

@@

Our analysis omits the space or time of execution for function calls. This information could provide additional insight for optimization. For example, spatial and temporal dimensions were shown to be important in the field of reduced precision. Future work could extent the depth of the analysis by including this information.
\MD{TODO Cite paper on spatial/temporal for RP}

@@

\section{Data Availability}
\label{sec:data-availability}
The entire OpenNeuro ds004513 v1.0.2 dataset is freely available to download at:
\\\href{https://openneuro.org/datasets/ds004513/versions/1.0.2}{https://openneuro.org/datasets/ds004513/versions/1.0.2}.
% TODO add compiled containers.
% TODO add csv files from profiled runs.
% The above can be distributed on DockerHub and Zenodo respectively.

\section{Code Availability}
\label{sec:code-availability}
The code to compile, profile the pipelines, and generate figures is publicly available at:
\\\href{https://github.com/mathdugre/mri-bottleneck}{https://github.com/mathdugre/mri-bottleneck}
% TODO Update URL to slashbin org, when created.

\section*{Acknowledgement}
% TODO

@@

\newpage
\onecolumn
\section*{Supplementary material}
\begin{figure*}
	\centering
	\includegraphics{figures/global-zipf_law.png}
	\caption{Distribution of the functions' CPU time compare to a Zipf's law distribution with $\alpha=1.8$. The y-axis shows the average CPU time for a function. The x-axis shows the percentage of functions ordered by decreasing CPU time. The data includes all functions from all pipelines.}
	\label{sup-fig:zips-law}
\end{figure*}

@@

\csvnames{csvcol}{2=\module, 3=\func, 4=\mean, 5=\std}
\csvstyle{hotspot}{
	tabular = |r|l|l|c|,
	table head = \hline & Module & Function & CPU Time (mean$\pm$std)\\\hline\hline,
	late after line = \\\hline,
	respect all,
	csvcol
}
\newcommand{\csvtable}[3]{
	\begin{table}[h]
		\resizebox*{\textwidth}{!}{
			\csvreader[hotspot]{#1}{}
			% \csvreader[hotspot, filter={\value{csvrow}<60}]{#1}{}
			{\thecsvrow & \module & \func & \tablenum[round-precision=2, round-mode=places]{\mean}$\pm$\tablenum[round-precision=2, round-mode=places]{\std}}
		}
		\caption{#2}
		\label{#3}
	\end{table}
}
